A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-06 podc_paper/introduction.tex
Rewording to save some space

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21} \here{try to find an older, more canonical reference}),
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For ease of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a tournament tree where each process has its own leaf.
Each process propagates its operations along the path from its leaf up to the root.
The root of the tree provides an ordering for all the operations,
and this ordering is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
To ensure wait-freedom, we use a helping mechanism:  a process propagating its operation 
to a node $n$ helps propagate all operations from both of $n$'s children.
If we explicitly stored all operations in the nodes, this helping would become very costly.
Instead, we use a novel implicit representation of collections
of operations that allows for quickly merging two collections from the two children of a node,
and for quickly accessing any particular operation in the collection.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

>>>>

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21} \here{try to find an older, more canonical reference}),
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first with step complexity that is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
For ease of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a tournament tree where each process has its own leaf.\here{Is "tournament tree" the right word here?  Maybe call it an ordering tree?}
Each process propagates its operations along the path from its leaf up to the root.
Operations in the root are ordered, 
and this order is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
Helping  ensures wait-freedom and avoids the CAS retry problem:  a process propagating its operation 
to a node $v$ helps propagate all operations from both of $v$'s children.
Explicitly storing all operations in the nodes would be too costly.
Instead, we use a novel implicit representation of sets
of operations that allows us to quickly merge two sets from the children of a node,
and quickly access any  operation in a~set.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2023-01-06 podc_paper/related.tex
Rewording to save some space

A number of papers have described ways to reduce contention in the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty or when concurrent operations 
can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
attempts to reduce contention by grouping concurrent enqueues into baskets.
An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
Enqueues within a basket determine their order among themselves without having to access the back of the queue.
However, if $p$ concurrent enqueues are in the same basket
the CAS retry problem occurs when they order themselves using CAS instructions.
Both modifications still have $\Omega(p)$ amortized step complexity.

>>>>

Many papers have described ways to reduce contention in the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty.
% or when concurrent operations can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
attempts to reduce contention by grouping concurrent enqueues into baskets.
An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
Enqueues within a basket determine their order among themselves without having to access the back of the queue.
However, if $p$ concurrent enqueues are in the same basket
the CAS retry problem occurs when they order themselves using CAS instructions.
Both modifications still have $\Omega(p)$ amortized step complexity.
